---
title: "illumina EPICv2 iScan Demo [Debug]"
author: Yuan Tian (champ450K@gmail.com)
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

This is a run markdown for ChAMP supporting on illumina HumanMethylation EPICv2 array. To replicate this run, you need to install the ChAMPdata (>=2.31.1), and ChAMP version >= 2.22.1. The latest ChAMPdata can be obtained [here](https://github.com/YuanTian1991/ChAMPdata). And the latest ChAMP is [here](https://github.com/YuanTian1991/ChAMP).

In this demo, I am using the defatul [demo data provided by illumina](https://emea.support.illumina.com/array/array_kits/infinium-methylationepic-beadchip-kit/downloads.html).

```{r, echo = TRUE, eval = FALSE}
library("ChAMP")
```

After the above library loading, you can check library version with `sessionInfo()`.

## 1. Import IDAT Data

Like previous version, ChAMP requires only one CSV to be put in a folder, along with IDAT files to be loaded.

```{r, echo = TRUE, eval = FALSE}
library("illuminaio")

source("~/Personal/ChAMP-Dev/ChAMP/R/champ.import.R")

myImport <- champ.import("./DemoDataEPICv2/", arraytype = "EPICv2") # This can also be set just as `EPIC`
```

## 2. Filtering

After loading, filtering will be done, specifically, the SNP mask is using the one provided by Zhou [here](http://zwdzwd.github.io/InfiniumAnnotation).

```{r, echo = TRUE, eval = FALSE}
library("impute")

source("~/Personal/ChAMP-Dev/ChAMP/R/champ.filter.R")
myFilter <- champ.filter(beta=myImport$beta, 
                         pd=myImport$pd, 
                         detP=myImport$detP,
                         beadcount=myImport$beadcount, 
                         ProbeCutoff=0.1,
                         arraytype = "EPICv2") # This can also be set just as `EPIC`


```

Below example shows if you need to do population-specific filtering. Currently, EPICv2 annotation only support AFR, AMR, EAS, EUR and SAS.

```{r, echo = TRUE, eval = FALSE}
myFilter <- champ.filter(beta=myImport$beta, 
                         pd=myImport$pd, 
                         detP=myImport$detP,
                         beadcount=myImport$beadcount,
                         population="EUR",
                         arraytype = "EPICv2") # This can also be set just as `EPIC`
```

## 3. Loading

Above two step can be merged into just one `champ.load()`. The defaul champ.load have two method: `ChAMP` or `minfi`, currently minfi package is not working for EPICv2, so only `ChAMP` works here.

```{r, echo = TRUE, eval = FALSE}
source("~/Personal/ChAMP-Dev/ChAMP/R/champ.load.R")

myLoad <- champ.load("./DemoDataEPICv2/", arraytype = "EPICv2")
```


## 4. CpG.GUI

Then, we can use CpG.GUI() to explore a beta matrix, along with a vector of phenotype.

```{r, echo = TRUE, eval = FALSE}
library("shiny")
library("shinythemes")
library("plotly")

source("~/Personal/ChAMP-Dev/ChAMP/R/CpG.GUI.R")

CpG.GUI(rownames(myLoad$beta), arraytype = "EPICv2")
```

## 5. QC

QC.GUI() can be used for check beta matrix

```{r, echo = TRUE, eval = FALSE}
library("shiny")
library("shinythemes")
library("plotly")
library("RColorBrewer")
library("dendextend")
library(isva)

source("~/Personal/ChAMP-Dev/ChAMP/R/EstDimRMTv2.R")
source("~/Personal/ChAMP-Dev/ChAMP/R/QC.GUI.R")

QC.GUI(beta=myLoad$beta, pheno = myLoad$pd$Sample_Group, arraytype = "EPICv2")
```

Also, champ.QC() function works, but I mostly only use QC.GUI() above.

```{r, echo = TRUE, eval = FALSE}
source("~/Personal/ChAMP-Dev/ChAMP/R/champ.QC.R")

champ.QC(beta = myLoad$beta, pheno = myLoad$pd$Sample_Group)
```

## 5. Normalisation

Since minfi is not working now, currently, champ.norm() only support `BMIQ` and `PBC` method. `SWAN` and `FunctionalNormalisation` is not working for now.

```{r, echo = TRUE, eval = FALSE}
library("doParallel")

source("~/Personal/ChAMP-Dev/ChAMP/R/champ.norm.R")
source("~/Personal/ChAMP-Dev/ChAMP/R/champ.BMIQ.R")


# I recommand BMIQ, which ONLY works for beta matrix.
myNorm <- champ.norm(myLoad$beta, method = "BMIQ", arraytype = "EPICv2")

# or PBC
source("~/Personal/ChAMP-Dev/ChAMP/R/DoPBC.R")
myNorm <- champ.norm(myLoad$beta, method = "PBC", arraytype = "EPICv2")
```

Below is the normalised data from QC.GUI()

```{r, echo = TRUE, eval = FALSE}
QC.GUI(beta=myNorm, pheno = myLoad$pd$Sample_Group, arraytype = "EPICv2")
```


## 6. SVD Check

Than champ.SVD() should be used to check confounding effect.

```{r, echo = TRUE, eval = FALSE}
source("~/Personal/ChAMP-Dev/ChAMP/R/champ.SVD.R")

champ.SVD(myNorm, pd = myLoad$pd)
```

The result shows there are confounding effect for Array, so I want to adjust it with Combat.

## 7. Combat Batch Effect Adjustment

Below code works for batch correction, but I do admit that current champ.runCombat() need to improve and failed constantly. A better method (maybe scBatch?) need to be incoprated.

```{r, echo = TRUE, eval = FALSE}
library("sva")

source("~/Personal/ChAMP-Dev/ChAMP/R/champ.runCombat.R")

myCombat <- champ.runCombat(beta=myNorm,pd=myLoad$pd,batchname=c("Array"))
```

## 7. DMP and GUI

Finally, below is the EWAS analysis, firstly do DMP analysis.

```{r, echo = TRUE, eval = FALSE}
library("limma")
source("~/Personal/ChAMP-Dev/ChAMP/R/champ.DMP.R")

# Below code exist because the origin sample group contains space, replace it with `_` works.
tmp_pheno <- myLoad$pd$Sample_Group
tmp_pheno <- gsub(" ", "_", tmp_pheno)

myDMP <- champ.DMP(beta = myNorm,pheno=tmp_pheno, arraytype = "EPICv2")
```

Then, CpG.GUI() can be used to check the DMP result:

```{r, echo = TRUE, eval = FALSE}
library("shiny")
library("shinythemes")
library("plotly")

source("~/Personal/ChAMP-Dev/ChAMP/R/DMP.GUI.R")

DMP.GUI(DMP=myDMP[[1]], beta=myNorm, pheno=myLoad$pd$Sample_Group)
```

It works nicely.

## 7. DMR and GUI

This is a challenge part. Firstly I believe `ProbeLasso` and `DMRcate` is not working, the formal requires a special king of old annotation, which I have not generated (but I don't want to use the old format anymore, so I will change the code to fit new annotation). DMRcate package can't be install on my computer...Like some user complained, I don't know why yet.

So, the `Bumphunter` is the only choice here for DMR. Actually this is the one I recommand.

```{r, echo = TRUE, eval = FALSE}
library("bumphunter")

source("~/Personal/ChAMP-Dev/ChAMP/R/champ.DMR.R")
tmp_pheno <- myLoad$pd$Sample_Group
tmp_pheno <- gsub(" ", "_", tmp_pheno)

myDMR <- champ.DMR(beta=myNorm, 
                   pheno=tmp_pheno,
                   method="Bumphunter", # Don't use othe method, only Bumphunter works.
                   arraytype = "EPICv2", 
                   cores=10)
```

Below is the DMP.GUI

```{r, echo = TRUE, eval = FALSE}
library("shiny")
library("shinythemes")
library("plotly")

source("~/Personal/ChAMP-Dev/ChAMP/R/DMR.GUI.R")

DMR.GUI(DMR=myDMR, beta=myNorm, pheno=tmp_pheno, arraytype = "EPICv2")
```

It works well, just a bit slow, I should accelerate it a bit with data.table....

## 7. Block and GUI

Similar as DMR, Block should works:

```{r, echo = TRUE, eval = FALSE}
source("~/Personal/ChAMP-Dev/ChAMP/R/champ.Block.R")
tmp_pheno <- myLoad$pd$Sample_Group
tmp_pheno <- gsub(" ", "_", tmp_pheno)

myBlock <- champ.Block(beta=myNorm, pheno = tmp_pheno, arraytype = "EPICv2")
```

Block.GUI is not working for now.

## 8. GSEA

The GSEA will take DMP and DMR result with various GO enrichment method. Neither `gometh` nor `ebayes` method currently is not working, only `fisher` works.


```{r, echo = TRUE, eval = FALSE}
library("missMethyl")
library("globaltest")

source("~/Personal/ChAMP-Dev/ChAMP/R/champ.GSEA.R")
source("~/Personal/ChAMP-Dev/ChAMP/R/champ.ebGSEA.R")

myGSEA <- champ.GSEA(beta=myNorm,
                     DMP=myDMP[[1]], 
                     DMR=myDMR,
                     pheno=tmp_pheno,
                     arraytype="EPICv2",
                     adjPval=0.05, 
                     method="fisher")


```


gometh is not supporting because it's a method from missMethyl package, which is also outdated seems. ebGSEA does not support because globaltest failed, I don't know why yet...

# 9. Refbase Cell Type Adjustment

Refbase function works for EPICv2.

```{r, echo = TRUE, eval = FALSE}
library("quadprog")
source("~/Personal/ChAMP-Dev/ChAMP/R/champ.refbase.R")

myRefBase <- champ.refbase(beta=as.data.frame(myNorm) ,arraytype="EPICv2")
```


# 10. CNA

This function need to improve a bit, which is not working now.

